---
title: "Classification of high/low saliva/gut transmission"
author: "Jakob Wirbel"
date: "2018-12-22"
output: BiocStyle::html_document
---


# Preparations 

```{r options}
options(stringsAsFactors = FALSE)
## beware the defaults in R!
```

```{r packages, message=FALSE}
library("SIAMCAT") # IMPORTANT: version 1.0.1 #
library("tidyverse")
library("readxl")
```

## Load Data

```{r load_data}
# features
feat <- read.table(file = './oral_gut.specI_abd_filtered.tsv',
                   sep='\t', header=TRUE, row.names = 1, 
                   check.names = FALSE)

# meta-data
meta <- read.table(file='./oral_gut.transmission_abd.tsv',
                   sep='\t', header=TRUE,
                   check.names = FALSE)

stopifnot(all(meta$sample.oral %in% colnames(feat)))
stopifnot(all(meta$sample.gut %in% colnames(feat)))

# taxa information
tax.info <- read_excel('./Table_S2.taxa_metadata.xlsx', sheet=2)
stopifnot(all(as.character(tax.info$Tax_ID) == rownames(feat)))

# rename features
rownames(feat) <- tax.info$Scientific_Name
```


## Preprocess transmission rates into quantiles

```{r preprocess_label}
percentiles_split <- c(0, 0.25, 0.75, 1)

# transform transmission rates into high/medium/low
meta <- meta %>%
  mutate(abd.saliva.factor = 
           cut(abd.saliva.transmitted, 
               breaks=quantile(abd.saliva.transmitted, probs=percentiles_split), 
               labels = c('low', 'medium', 'high'),
               include.lowest = TRUE)) %>%
  mutate(abd.stool.factor = 
           cut(abd.stool.transmitted, 
               breaks=quantile(abd.stool.transmitted, probs=percentiles_split), 
               labels=c('low', 'medium', 'high'),
               include.lowest = TRUE))

# plot histograms
# saliva
meta %>%
  ggplot(aes(x=reorder(sample.oral, -abd.saliva.transmitted), 
             y=abd.saliva.transmitted, fill=abd.saliva.factor)) +
    geom_bar(stat='identity', col=NA) + 
    theme_classic() + 
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
    xlab('Samples') + 
    ylab('Saliva transmission rate') + 
    scale_fill_manual(values=c('#d9544d', '#39ad48', '#3b5b92'), name='Group') +
    scale_y_continuous(limits=c(0, 0.8), expand = c(0, 0))
# stool
meta %>%
  ggplot(aes(x=reorder(sample.gut, -abd.stool.transmitted), 
             y=abd.stool.transmitted, fill=as.factor(abd.stool.factor))) +
    geom_bar(stat='identity', col=NA) + 
    theme_classic() + 
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
    xlab('Samples') + 
    ylab('Saliva transmission rate') + 
    scale_fill_manual(values=c('#d9544d', '#39ad48', '#3b5b92'), name='Group') +
    scale_y_continuous(limits=c(0, 0.4), expand = c(0, 0))
```


## Create SIAMCAT label 

Now, we create the label objects for the different classification tasks we 
want to investigate.

```{r create_label}

# delete superfluous metadata (not needed for plotting later)
meta <- meta %>% 
  select(sample.oral, sample.gut, group, abd.saliva.transmitted,
         abd.stool.transmitted, abd.saliva.factor, abd.stool.factor)

# prediction using the saliva samples as input features
meta.saliva <- meta
rownames(meta.saliva) <- meta$sample.saliva # rownames are needed for SIAMCAT
meta.saliva <- meta.saliva[,-c(1,2)]

# predict saliva transmission rates
label.saliva.saliva <- create.label(
  'abd.saliva.factor', case='high', control='low', meta=meta.saliva
  )
# predict gut transmission rates
label.saliva.stool <- create.label(
  'abd.stool.factor', case='high', control='low', meta=meta.saliva
)


# and the same for gut samples
meta.stool <- meta
meta.stool <- meta.stool[,-c(1,2)]
# prediction using the gut samples as input features
rownames(meta.stool) <- meta$sample.gut
# predict saliva transmission rates
label.stool.saliva <- create.label(
  'abd.saliva.factor', case='high', control='low', meta=meta.stool
  )
# predict gut transmission rates
label.stool.stool <- create.label(
  'abd.stool.factor', case='high', control='low', meta=meta.stool
)
```

# Machine learning part

## Machine learning pipeline

Since we will train several models with the same parameters, we can create a 
function to train models.

```{r ml_pipeline}
ml.pipeline <- function(feat, meta, label){
  siamcat <- siamcat(feat=feat, meta=meta, label=label)
  siamcat <- normalize.features(siamcat, norm.method = 'log.std', 
                                feature.type = 'original')
  siamcat <- create.data.split(siamcat, num.folds = 10, num.resample = 10)
  siamcat <- train.model(siamcat, method='lasso')
  siamcat <- make.predictions(siamcat)
  siamcat <- evaluate.predictions(siamcat)
  return(siamcat)
}
```

## Train models

```{r train_models, results='hide', message=FALSE}
if (file.exists('./siamcat_models.RData')){
  load('./siamcat_models.RData')
} else {
  siamcat.saliva.stool <- ml.pipeline(feat=feat, 
                                    meta=meta.saliva, 
                                    label=label.saliva.stool)
  siamcat.saliva.saliva <- ml.pipeline(feat=feat,
                                   meta=meta.saliva,
                                   label=label.saliva.saliva)
  siamcat.stool.saliva <- ml.pipeline(feat=feat,
                                    meta=meta.stool,
                                    label=label.stool.saliva)
  siamcat.stool.stool <- ml.pipeline(feat=feat,
                                     meta=meta.stool,
                                     label=label.stool.stool)
  save(siamcat.stool.stool, siamcat.stool.saliva, 
       siamcat.saliva.saliva, siamcat.saliva.stool,
       file = './siamcat_models.RData')
}
```

## Evaluation

```{r eval_plot}
model.evaluation.plot('stool-stool'=siamcat.stool.stool,
                      'stool-saliva'=siamcat.stool.saliva,
                      'saliva-saliva'=siamcat.saliva.saliva,
                      'saliva-stool'=siamcat.saliva.stool)
```

```{r model_interpretation_plot}
model.interpretation.plot(siamcat.saliva.saliva,
                          fn.plot='./interpretation_saliva_saliva.pdf')
model.interpretation.plot(siamcat.saliva.stool,
                          fn.plot='./interpretation_saliva_stool.pdf')
model.interpretation.plot(siamcat.stool.saliva,
                          fn.plot='./interpretation_stool_saliva.pdf')
model.interpretation.plot(siamcat.stool.stool,
                          fn.plot='./interpretation_stool_stool.pdf')
```

Let's have a look at the features carrying the most model weight.

```{r plot_model_importances_vs_transmission}
df.plot <- tibble(
  tax_ID = tax.info$Scientific_Name,
  `stool-stool` = feature_weights(siamcat.stool.stool)
    [as.character(tax.info$Scientific_Name),"mean.rel.weight"],
  `stool-saliva` = feature_weights(siamcat.stool.saliva)
    [as.character(tax.info$Scientific_Name),"mean.rel.weight"],
  `saliva-stool` = feature_weights(siamcat.saliva.stool)
    [as.character(tax.info$Scientific_Name),"mean.rel.weight"],
  `saliva-saliva` = feature_weights(siamcat.saliva.saliva)
    [as.character(tax.info$Scientific_Name),"mean.rel.weight"],
  category = tax.info$transmitter.category,
  transmission = as.numeric(tax.info$transmission_score.median))

df.plot <- df.plot %>% 
  gather(value=weight, key=model, -category, -transmission, -tax_ID)

df.plot %>% 
  ggplot(aes(x=category, y=weight, fill=model)) +
    geom_boxplot()

df.plot %>% 
  select(model, category, weight) %>% 
  group_by(model, category) %>% 
  summarize(n_nonzero=sum(weight!=0, na.rm=TRUE)) %>% 
  ggplot(aes(x=model, y=n_nonzero, fill=category)) +
    geom_bar(stat='identity', position = position_dodge())

df.plot %>% 
  ggplot(aes(y=transmission, x=weight, col=category)) +
    geom_point() + 
    facet_grid(model~.)

```


## Eliminate transmitter species

Let's eliminate the transmitter species and train some models again.

```{r eliminate_features, results='hide', message=FALSE}
if (file.exists('./siamcat_models_eliminated.RData')){
  load('./siamcat_models_eliminated.RData')
} else {
  # train without transmitters
  feat.red.1 <- feat[
    tax.info %>% 
      filter(transmitter.category != 'transmitter') %>%
      pull(Scientific_Name) %>% 
      as.character,]
  
  siamcat.stool.wo.trans <- ml.pipeline(feat=feat.red.1, 
                                        meta=meta.stool, 
                                        label=label.stool.stool)
  
  siamcat.saliva.wo.trans <- ml.pipeline(feat=feat.red.1, 
                                         meta=meta.stool, 
                                         label=label.stool.saliva)
  save(siamcat.stool.wo.trans, siamcat.saliva.wo.trans, 
       file = './siamcat_models_eliminated.RData')
}
```

```{r model_plot_again}
model.evaluation.plot(siamcat.stool.wo.trans, siamcat.saliva.wo.trans)

model.interpretation.plot(siamcat.stool.wo.trans,
                          fn.plot = './interpretation_stool_eleminated_1.pdf')

model.interpretation.plot(siamcat.saliva.wo.trans,
                          fn.plot = './interpretation_saliva_eleminated_1.pdf')
```

## Combined Evaluation Plot

Now that we have all the models trained, we can compare how well the different
features can predict the transmission group in saliva and stool.

```{r grouped_evalation}
# for saliva samples
model.evaluation.plot('based on saliva'=siamcat.saliva.saliva,
                      'based on stool'=siamcat.stool.saliva,
                      'based on stool\n(without transmitter)'=
                        siamcat.saliva.wo.trans)

model.evaluation.plot('based on saliva'=siamcat.saliva.stool,
                      'based on stool'=siamcat.stool.stool,
                      'based on stool\n(without transmitter)'=
                        siamcat.stool.wo.trans)
```

# SessionInfo

```{r session_info}
sessionInfo()
```
